---
title: "SF Crime Classification"
author: "Jerome & Chris"
date: "6 September 2015"
output: html_document
---

# Introduction
This is an easy way to share our exploratory data analysis for the Kaggle Competition - https://www.kaggle.com/c/sf-crime.

# Data Load
Normally I'd load the data in from the website in the script, but it's not working so I'll just assume that it's in the working directory.

Note: I experimented with setting StringsAsFactors = FALSE, but I ran into some crashes, so am sticking with the default.

```{r DataLoad, cache = TRUE}
train <- read.csv("train.csv")
#test <- read.csv("test.csv")
```

# Data Cleaning
Description and Resolution fields do not exist in the test dataset. Leave them in for exploration to see whether we can find any patterns in the remaining fields which could be good features. We'll remove them from dataset before fitting a model.

- There are 67 records where Latitude (Y) = 90, which is the North pole. The summary below shows that these are evenly distributed, so they can be removed.

```{r Latitude90}
summary(train[train$Y > 40,])

train2 <- train[train$Y < 40,] 
```


# Feature Engineering
## Time and Date
Lets split the date into three features - Year, Month and Hour. Then drop the date field.

One possible alternative would be TimeOfDay. Splitting hour into six hour segments:

- Midnight - 6am: Overnight
- 6am - Midday: Morning
- Midday - 6pm: Afternoon
- 6pm - Midnight: Evening

```{r YearMonthHour}
library(lubridate)
train2$Year <- year(train2$Dates)
train2$Month <- month(train2$Dates)
train2$Hour <- hour(train2$Dates)
```

## Address
One simple approach would be to categorise addresses by Block or Corner. Output below shows that all addresses are either block or corner.

```{r Address}
train2$Block <- grepl("Block of", train2$Address)

# Check that all other rows are corners
block <- sum(train2$Block)
corner <- sum(grepl("/", train2$Address))
block + corner
nrow(train2)
```


## Subset of data
To enable quick plotting of trends, create a subset of the data, removing the rows which are not in the model.

```{r Subset}
set.seed(42)
sample.rows <- sample(1:nrow(train2), 5000)
train3 <- train2[sample.rows,]
train3$Dates <- NULL
train3$Descript <- NULL
train3$Resolution <- NULL
train3$Address <- NULL
train3$X <- NULL
train3$Y <- NULL
```

# Exploratory Data Analysis
To begin with we'll create summary dataframes and charts for the proportion of each Category, by the main address and date-time factors. We'll calculate a ratio to highlight trends in the data.

## Incident Category by Hour
Inspection of the plots below indicate that splitting the Hour into four, six hour periods would generate a good approximation to the data. Crime categories tend to have a profile that is greater during night, morning, afternoon and evening.

```{r CategoryByHour, fig.width = 10, fig.height = 8}
suppressMessages(library(dplyr))
library(ggplot2)
ratio.Hour <- train2 %>%
  group_by(Category, Hour) %>%
  summarize(count = n()) %>%
  group_by(Hour) %>%
  mutate (hc = sum(count)) %>%
  ungroup() %>%
  mutate (ratio = count / hc)

ggplot(ratio.Hour, aes(Hour, ratio)) + 
  geom_line(aes(colour = Category)) + 
  facet_wrap(~ Category, ncol = 5, scales = "free_y") +
  scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) +
  theme(legend.position="none")
```

## Incident Category by Hour
Inspection of the plots below indicate that we should retain Day as a factor with seven levels. The obvious alternative is a two level factor with "Weekday" and "Weekend", but there is a lot of variation in the data which would be lost if we went with this approach.

```{r CategoryByDay, fig.width = 10, fig.height = 8}
ratio.Day <- train2 %>%
  group_by(Category, DayOfWeek) %>%
  summarize(count = n()) %>%
  group_by(DayOfWeek) %>%
  mutate(dc = sum(count)) %>%
  ungroup() %>%
  mutate(ratio = count / dc) 

ratio.Day$DayOfWeek <- factor(ratio.Day$DayOfWeek,
                              levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday","Saturday", "Sunday"),
                              labels = c("M", "Tu", "W", "Th", "F", "Sa", "Su"),
                              ordered = TRUE)

ggplot(ratio.Day, aes(DayOfWeek, ratio)) + 
  geom_line(aes(colour = Category, group = Category)) + 
  facet_wrap(~ Category, ncol = 5, scales = "free_y") +
  theme(legend.position="none")
```