---
title: "SF Crime Exploratory Analysis"
author: "Jerome & Chris"
date: "6 September 2015"
output: html_document
---

# Introduction
This is an easy way to share our exploratory data analysis for the Kaggle Competition - https://www.kaggle.com/c/sf-crime.

# Data Load
Normally I'd load the data in from the website in the script, but it's not working so I'll just assume that it's in the working directory.

Note: I experimented with setting StringsAsFactors = FALSE, but I ran into some crashes, so am sticking with the default.

```{r DataLoad, cache = TRUE}
train <- read.csv("train.csv")
#test <- read.csv("test.csv")
```

# Data Cleaning
Description and Resolution fields do not exist in the test dataset. Leave them in for exploration to see whether we can find any patterns in the remaining fields which could be good features. We'll remove them from dataset before fitting a model.

- There are 67 records where Latitude (Y) = 90, which is the North pole. The summary below shows that these are evenly distributed, so they can be removed.

```{r Latitude90, cache = TRUE}
summary(train[train$Y == 90,])

train2 <- train[train$Y < 90,] 
```


# Feature Engineering
## Time and Date
Lets split the date into three features - Year, Month and Hour. Then drop the date field.

One possible alternative would be TimeOfDay. Splitting hour into six hour segments:

- Midnight - 6am: Overnight
- 6am - Midday: Morning
- Midday - 6pm: Afternoon
- 6pm - Midnight: Evening

```{r YearMonthHour, cache = TRUE}
library(lubridate)
train2$Year <- year(train2$Dates)
train2$Month <- month(train2$Dates)
train2$Hour <- hour(train2$Dates)
```

## Address
One simple approach would be to categorise addresses by Block or Corner. Output below shows that all addresses are either block or corner.

```{r Address, cache = TRUE}
train2$Block <- grepl("Block of", train2$Address)

# Check that all other rows are corners
block <- sum(train2$Block)
corner <- sum(grepl("/", train2$Address))
block + corner
nrow(train2)
```


## Subset of data
To enable quick plotting of trends or experimentation with different models, create a subset of the data, removing the rows which we won't use in the model.

```{r Subset, cache = TRUE}
set.seed(42)
sample.rows <- sample(1:nrow(train2), 5000)
train3 <- train2[sample.rows,]
train3$Dates <- NULL
train3$Descript <- NULL
train3$Resolution <- NULL
train3$Address <- NULL

```

# Exploratory Data Analysis
To begin with we'll order the categories, with the most frequent first. Then create summary dataframes and charts for the proportion of each category, by each address and date-time factor. We'll calculate a ratio to highlight trends in the data.

## Incident Category
```{r Category, cache = TRUE, fig.height = 10}
suppressMessages(library(dplyr))
library(ggplot2)

total.category <- table(train2$Category)
order.category <- order(total.category, decreasing = TRUE)
train2$Category <- factor(train2$Category, levels = names(total.category[order.category]))

ggplot(train2, aes(Category)) +
  geom_histogram(aes(fill = Category)) +
  scale_y_continuous() +
  scale_x_discrete(limits = levels(train2$Category)[39:1]) +
  theme(legend.position="none", axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip()

```

## Incident Category by Hour
Inspection of the plots below shows that there is significant variation in the type of crime depending on the time. Therefore we will not group the time of day into larger blocks of time. 

```{r CategoryByHour, fig.width = 10, fig.height = 8, cache = TRUE}
ratio.Hour <- train2 %>%
  group_by(Category, Hour) %>%
  summarize(count = n()) %>%
  group_by(Hour) %>%
  mutate (hc = sum(count)) %>%
  ungroup() %>%
  mutate (ratio = count / hc)

ggplot(ratio.Hour, aes(Hour, ratio)) + 
  geom_line(aes(colour = Category)) + 
  facet_wrap(~ Category, ncol = 5, scales = "free_y") +
  scale_x_continuous(breaks=c(0, 6, 12, 18, 24)) +
  expand_limits(y = 0) + 
  theme(legend.position="none")
```

## Incident Category by Day
Inspection of the plots below indicate that there isn't as much variation as in Incident Category by Hour, but there can still be up to 25% variation in the top 15 crimes. To reduce the number of variables the obvious alternative is a two level factor with "Weekday" and "Weekend". This might be a reasonable approximation, but isn't clear cut. Let's start with a 7 level factor, and try a 2 level factor as a possible improvement to the model.


```{r CategoryByDay, fig.width = 10, fig.height = 8, cache = TRUE}
ratio.Day <- train2 %>%
  group_by(Category, DayOfWeek) %>%
  summarize(count = n()) %>%
  group_by(DayOfWeek) %>%
  mutate(dc = sum(count)) %>%
  ungroup() %>%
  mutate(ratio = count / dc) 

ratio.Day$DayOfWeek <- factor(ratio.Day$DayOfWeek,
                              levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday","Saturday", "Sunday"),
                              labels = c("M", "Tu", "W", "Th", "F", "Sa", "Su"),
                              ordered = TRUE)

ggplot(ratio.Day, aes(DayOfWeek, ratio)) + 
  geom_line(aes(colour = Category, group = Category)) + 
  facet_wrap(~ Category, ncol = 5, scales = "free_y") +
  expand_limits(y = 0) + 
  theme(legend.position="none")
```


## Incident Category by Month
The differences between the monthly totals are marginal for the top crimes. Suggest dropping this factor from the model.

```{r CategoryByMonth, fig.width = 10, fig.height = 8, cache = TRUE}
ratio.Month <- train2 %>%
  group_by(Category, Month) %>%
  summarize(count = n()) %>%
  group_by(Month) %>%
  mutate(dc = sum(count)) %>%
  ungroup() %>%
  mutate(ratio = count / dc) 

ggplot(ratio.Month, aes(Month, ratio)) + 
  geom_line(aes(colour = Category, group = Category)) + 
  facet_wrap(~ Category, ncol = 5, scales = "free_y") +
  scale_x_continuous(breaks=c(1, 4, 7, 10), 
                     minor_breaks = c(2, 3, 5, 6, 8, 9, 11, 12),
                     labels = c("Jan", "Apr", "Jul", "Oct")) +
  expand_limits(y = 0) + 
  theme(legend.position="none")
```


## Incident Category by Year
Inspection of the plots below indicate that year is an important factor for use in the model.

```{r CategoryByYear, fig.width = 10, fig.height = 8, cache = TRUE}
ratio.Year <- train2 %>%
  group_by(Category, Year) %>%
  summarize(count = n()) %>%
  group_by(Year) %>%
  mutate(dc = sum(count)) %>%
  ungroup() %>%
  mutate(ratio = count / dc) 

ggplot(ratio.Year, aes(Year, ratio)) + 
  geom_line(aes(colour = Category, group = Category)) + 
  facet_wrap(~ Category, ncol = 5, scales = "free_y") +
  scale_x_continuous(breaks=c(2003, 2007, 2011, 2015)) +  
  expand_limits(y = 0) + 
  theme(legend.position="none")
```


## Incident Category by PdDistrict
Inspection of the plots below indicate that PdDistrict is a valuable factor in the model.

```{r CategoryByPdDistrict, fig.width = 10, fig.height = 8, cache = TRUE}
ratio.PdDistrict <- train2 %>%
  group_by(Category, PdDistrict) %>%
  summarize(count = n()) %>%
  group_by(PdDistrict) %>%
  mutate(dc = sum(count)) %>%
  ungroup() %>%
  mutate(ratio = count / dc) 

ggplot(ratio.PdDistrict, aes(PdDistrict, ratio)) + 
  geom_line(aes(colour = Category, group = Category)) + 
  facet_wrap(~ Category, ncol = 5, scales = "free_y") +
  expand_limits(y = 0) + 
  theme(legend.position="none", axis.text.x = element_text(angle = 45, hjust = 1))
```


## Incident Category by Block
Inspection of the plots below indicate that there is a very significant difference between block and corner. This factor should be retained in the model.

```{r CategoryByBlock, fig.width = 10, fig.height = 8, cache = TRUE}
ratio.Block <- train2 %>%
  group_by(Category, Block) %>%
  summarize(count = n()) %>%
  group_by(Block) %>%
  mutate(dc = sum(count)) %>%
  ungroup() %>%
  mutate(ratio = count / dc) 

ggplot(ratio.Block, aes(Block, ratio)) + 
  geom_boxplot(aes(colour = Category)) + 
  facet_wrap(~ Category, ncol = 5, scales = "free_y") +
  scale_x_discrete(labels = c("Corner", "Block")) +
  expand_limits(y = 0) + 
  theme(legend.position="none")
```

